Forecasting the 2024 U.S. Presidential Election: Analyzing Poll Quality, Sample Size, and Regional Variations

### 2024-11-02 paper/paper.qmd
Updated QMD file to correct remaining issues

## Setup for workspace ##
# Load necessary libraries
# Load the tidyverse meta-package (includes dplyr, ggplot2, and readr)
library(tidyverse)
# Load the gt library for better visualization of some tables
library(gt)
# Load the knitr library for better visualization of some tables
library(knitr)
# The broom library is used for converting the model output into a tidy data frame
library(broom)
# this is used for presenting model output
library(modelsummary)
# For date handling
library(lubridate)

@@

# Introduction

@@

The paper is structured as follows: we begin by describing the dataset and its key variables, after which the model is presented, followed by the results and their implications. We use the statistical programming language R [@citeR] to conduct our analysis. In the last part of the paper, we critique the method itself and attempt to abstract some of the learnings into our understanding of the world.

# Data

@@

# Load the data
polls_data_cleaned <- read_csv(here::here("data/analysis_data/election_polls_cleaned.csv"))
# Summary statistics for numeric variables
# Basic statistics (Mean and SD)
basic_stats <- polls_data_cleaned %>%
  summarise(
    Poll_Score_Mean = mean(pollscore, na.rm = TRUE),
    Poll_Score_SD = sd(pollscore, na.rm = TRUE),
    Sample_Size_Mean = mean(sample_size, na.rm = TRUE),
    Sample_Size_SD = sd(sample_size, na.rm = TRUE),
    Pct_Mean = mean(pct, na.rm = TRUE),
    Pct_SD = sd(pct, na.rm = TRUE)
  )

# Min and Max statistics
min_max_stats <- polls_data_cleaned %>%
  summarise(
    Poll_Score_Min = min(pollscore, na.rm = TRUE),
    Poll_Score_Max = max(pollscore, na.rm = TRUE),
    Sample_Size_Min = min(sample_size, na.rm = TRUE),
    Sample_Size_Max = max(sample_size, na.rm = TRUE),
    Pct_Min = min(pct, na.rm = TRUE),
    Pct_Max = max(pct, na.rm = TRUE)
  )

# Frequency tables for categorical variables
categorical_summary <- polls_data_cleaned %>%
  select(state, party, candidate_name) %>%
  summarise(
    State_Count = n_distinct(state),
    Party_Count = n_distinct(party),
    Candidate_Count = n_distinct(candidate_name)
  )

# Display the basic summary statistics
basic_stats %>%
  kable(caption = "Basic Summary Statistics (Mean and SD)")

@@

# Model

## Model Specification and Justification:
The Generalized Linear Model (GLM) that we use in this analysis is a logistic regression model. It aims to predict the likelihood that a candidate wins a poll. The outcome variable we use is a binary one, win, with 1 indicating that a candidate won a poll (defined as receiving more than 50% of the vote), and 0 indicates that they did not.

@@

@@

Presented below are the significant variables from the analysis.

@@

# Fit the GLM model using logistic regression
glm_model <- glm(win ~ pollscore + sample_size + state + party_binary, family = binomial, data = polls_data_cleaned)

# Tidy the model output using broom's tidy() function
glm_tidy <- tidy(glm_model)

# Filter for statistically significant variables (p < 0.05)
glm_significant <- glm_tidy %>%
  filter(p.value < 0.05)

# Display the filtered results in a clean table format
glm_significant %>%
  kable(caption = "Statistically Significant Variables from the GLM Model")

@@

# Results

@@

## Visualizing Predictions

@@

# Predicted probability based on pollscore and sample_size
ggplot(polls_data_cleaned, aes(x = pollscore, y = predict(glm_model, type = "response"), color = sample_size)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Predicted Probability of Winning Based on Pollscore and Sample Size", x = "Poll Score", y = "Predicted Probability")
```

# Discussion

One key takeaway from this analysis is that polling quality significantly affects predictions. Polls with higher quality scores (i.e. those employing robust sampling methods, larger sample sizes, and transparent methodologies) are more likely to predict outcomes closer to the eventual election results. This highlights the importance of weighting polling data based on its quality when conducting election forecasts. The model confirms that using low-quality polls can distort predictions, especially in closely contested states, where small changes in the polling data can result in different outcomes.

Another important finding is the strong state-level effects. It is clear that political geography plays a role in influencing election outcomes within the United States context. This is seen in situations where nationwide polls indicate a tight race, the dynamics at the state level significantly influence the ultimate outcome of the Electoral College vote. States such as California or Texas that consistently align with one party highlight how state allegiance tends to be more indicative of electoral trends, than national polling data alone. Thus, election forecasting models that do not account for state-specific trends are likely to misestimate the actual results.

@@

@@

@@

- **MTurk**, **Prolific**, and **YouGov** panels to access diverse respondents.

@@

- **Duplicate response filtering** using IP addresses.
- **Post-stratification weighting** - this will be used to adjust the sample to match U.S. population proportions based on gender, race, and age.

@@

Results from the survey will be aggregated with other polls (e.g., **Gallup**, **Ipsos**, **YouGov**) using **weighted averages**. **Bayesian updating** will incorporate historical trends and account for shifts in voter behavior. The aggregated margin of error is projected at **±1%**, ensuring high forecast precision.

@@

9. **How important is the issue of the economy in your decision?**
   - Extremely important
   - Very important
   - Moderately important
   - Slightly important
   - Not important at all

10. **How favorable is your opinion of Kamala Harris?**
    - Very favorable
    - Somewhat favorable
    - Neutral
    - Somewhat unfavorable
    - Very unfavorable

11. **How favorable is your opinion of Donald Trump?**
    - Very favorable
    - Somewhat favorable
    - Neutral
    - Somewhat unfavorable
    - Very unfavorable

>>>>

## Setup for workspace ##
# Load necessary libraries
# Load the tidyverse meta-package (includes dplyr, ggplot2, and readr)
library(tidyverse)
# Load the gt library for better visualization of some tables
library(gt)
# The broom library is used for converting the model output into a tidy data frame
library(broom)
# use knitr for better table formatting with kable
library(knitr)

@@

# Introduction {#introduction}

@@

We start by describing the dataset in detail and discussing key variables in the Data section (see [Data](#data)), where we present summary statistics and visualizations to better understand the data. The Model section (see [Model](#model)) introduces the Generalized Linear Model (GLM). This includes an exaplanation of the mathematical framework, assumptions, and variable selection process. We then detail the Results (see [Results](#results)), where we highlight key findings on how poll characteristics and state-level differences impact election forecasts. Finally, the Discussion (see [Discussion](#discussion)) addresses the implications of our findings, the limitations of our approach, and suggestions for future research. In the last part of the paper, we critique the method itself and attempt to abstract some of the learnings into our understanding of the world.

We use the statistical programming language R [@citeR] to conduct our analysis, the tidyverse package for data manipulation and visualization [@tidyverse] and tables were created using the gt package [@gt]. We used the knitr package to format tables and generate a dynamic report [@knitr]. Finally, model summaries were tidied using the broom package [@broom].

# Data {#data}

@@

# Load the data
polls_data_cleaned <- read_csv(here::here("data/analysis_data/election_polls_cleaned.csv"))
# Summary statistics for numeric variables

# Basic statistics (Mean and SD)
basic_stats <- polls_data_cleaned %>%
  summarise(
    Poll_Score_Mean = round(mean(pollscore, na.rm = TRUE), 2),
    Poll_Score_SD = round(sd(pollscore, na.rm = TRUE), 2),
    Sample_Size_Mean = round(mean(sample_size, na.rm = TRUE), 2),
    Sample_Size_SD = round(sd(sample_size, na.rm = TRUE), 2),
    Pct_Mean = round(mean(pct, na.rm = TRUE), 2),
    Pct_SD = round(sd(pct, na.rm = TRUE), 2)
  ) %>%
  rename(
    "Average Poll Score" = Poll_Score_Mean,
    "Poll Score Standard Deviation" = Poll_Score_SD,
    "Average Sample Size" = Sample_Size_Mean,
    "Sample Size Standard Deviation" = Sample_Size_SD,
    "Average Percentage" = Pct_Mean,
    "Percentage Standard Deviation" = Pct_SD
  )

# Min and Max statistics
min_max_stats <- polls_data_cleaned %>%
  summarise(
    Poll_Score_Min = round(min(pollscore, na.rm = TRUE), 2),
    Poll_Score_Max = round(max(pollscore, na.rm = TRUE), 2),
    Sample_Size_Min = round(min(sample_size, na.rm = TRUE), 2),
    Sample_Size_Max = round(max(sample_size, na.rm = TRUE), 2),
    Pct_Min = round(min(pct, na.rm = TRUE), 2),
    Pct_Max = round(max(pct, na.rm = TRUE), 2)
  ) %>%
  rename(
    "Minimum Poll Score" = Poll_Score_Min,
    "Maximum Poll Score" = Poll_Score_Max,
    "Minimum Sample Size" = Sample_Size_Min,
    "Maximum Sample Size" = Sample_Size_Max,
    "Minimum Percentage" = Pct_Min,
    "Maximum Percentage" = Pct_Max
  )

# Frequency tables for categorical variables
categorical_summary <- polls_data_cleaned %>%
  select(state, party, candidate_name) %>%
  summarise(
    State_Count = n_distinct(state),
    Party_Count = n_distinct(party),
    Candidate_Count = n_distinct(candidate_name)
  ) %>%
  rename(
    "Number of States" = State_Count,
    "Number of Parties" = Party_Count,
    "Number of Candidates" = Candidate_Count
  )
# Display the basic summary statistics
basic_stats %>%
  kable(caption = "Basic Summary Statistics (Mean and SD)")

@@

# Model {#model}

## Model Specification and Justification
The Generalized Linear Model (GLM) that we use in this analysis is a logistic regression model. It aims to predict the likelihood that a candidate wins a poll. The outcome variable we use is a binary one, win, with 1 indicating that a candidate won a poll (defined as receiving more than 50% of the vote), and 0 indicates that they did not.

@@

5. **End Date (`end_date`)**: This variable represents the date when the poll concluded. Including this variable in the model accounts for temporal trends, as public opinion can shift significantly over time. The end_date variable helps capture how voter sentiment evolves as the election approaches, providing a way to model these changes.

@@

Presented below are the significant variables from the analysis, capturing the different battleground states.

@@

# Fit the GLM model using logistic regression
glm_model <- glm(
  win ~ pollscore + sample_size + state + party_binary + end_date,
  family = binomial,
  data = polls_data_cleaned
)

# Use broom to tidy the model
glm_summary <- broom::tidy(glm_model)
# Filter for statistically significant variables (e.g., p-value < 0.05)
significant_vars <- glm_summary %>%
  filter(p.value < 0.05)

# Remove p-values from the table
significant_vars <- significant_vars %>% select(term, estimate, std.error)

significant_vars %>%
  kable(
    caption = "Estimated coefficients from the Generalized Linear Model, showing the effects of poll quality, sample size, state, and party affiliation on the probability of a candidate winning a poll. Only statistically significant variables are included to highlight key predictors."
  )

@@

# Results {#results}

@@

##  Impact of Poll Quality and Sample Size on Predictions

@@

# Predicted probability based on pollscore and sample_size
ggplot(polls_data_cleaned, aes(x = pollscore, y = predict(glm_model, type = "response"), color = sample_size)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Predicted probabilities of a candidate winning a poll as a function of poll quality (pollscore) and sample size. The plot demonstrates how higher poll quality scores and larger sample sizes are associated with more accurate and moderated predictions, particularly in battleground states", x = "Poll Score", y = "Predicted Probability")
```

# Discussion - Key Influences on 2024 Election Predictions {#discussion}

## Key Findings

The conclusion from the analysis is that both poll quality and sample size predict the probability of a candidate winning a poll. Quality polls with higher pollscore values generally make more conservative and balanced predictions - especially in battleground states. For instance, in Florida, Trump trails the pack 52% to 51% in higher-quality polls. In Pennsylvania, Harris' margin is also slimmer at 52%. This demonstrates how poll reliability may influence impressions of candidate support.

Sample size also matters. More large samples result in predictions with less variability. The model demonstrates that smaller polls - especially those with fewer than 1,000 respondents - tend to exaggerate support for one candidate. Larger polls - those with more than 5,000 respondents - moderate these extremes and reflect voter sentiment better. In Arizona, for example, larger, more robust samples drop Trump's predicted support by 2% to 49%.

## State-Level Effects

The electoral landscape requires understanding state-specific dynamics. The model shows that voter preferences vary widely across states, with Trump strongly favored in Republican-leaning Texas and Florida and Harris well backed in Democratic strongholds California and New York. And the races are tighter in Ohio and Pennsylvania - another example of how state analysis is critical to a broader election forecast.

Small but significant voter differences exist in battleground states like Florida and Arizona. The results indicate that slight shifts in voter sentiment or polling methodology may change predicted outcomes significantly in these key states. All this points to the need for more frequent, high-quality polling in swing states that capture voter behaviour.

## Limitations and Assumptions

@@

## Future Directions

@@

**Justification**: Stratified random sampling addresses heterogeneity in voter behavior and ensures that our sample is representative of the voting-age population. That approach is especially useful in capturing variation within battlefield states where minor variations in voter turnout could tip the election in a big way [@gelman2007].

@@

- **MTurk**, **Prolific**, and **YouGov** panels to access diverse respondents.Studies show that online recruitment platforms may produce quality survey data when sampling and weighting are done properly [@huff2015].

@@

- **Duplicate response filtering** using IP addresses.
- **Post-stratification weighting** - this will be used to adjust the sample to match U.S. population proportions based on gender, race, and age [@pasek2015].

@@

Results from the survey will be aggregated with other polls (e.g., **Gallup**, **Ipsos**, **YouGov**) using **weighted averages**. **Bayesian updating** will incorporate historical trends and account for shifts in voter behavior [@jackman2009]. The aggregated margin of error is projected at **±1%**, ensuring high forecast precision.

@@

9. **How important is the issue of the economy in your decision?** 
   - Please describe why the economy is important to your voting decision:

10. **Rank the following issues in order of importance for your voting decision** 
   - Economy
   - Healthcare
   - Immigration
   - Climate Change
   - Education

11. **How favorable is your opinion of Kamala Harris?** 
    - On a scale of 1 to 10, where 1 is very unfavorable and 10 is very favorable, please rate your opinion of Kamala Harris:

12. **How favorable is your opinion of Donald Trump?** 
    - On a scale of 1 to 10, where 1 is very unfavorable and 10 is very favorable, please rate your opinion of Donald Trump:

13. **How closely do you follow election news?** 
    - Please explain how often and through which sources you follow election news:

14. **What factors most influence your voting decision?**
    - Please provide details about the factors that are most important to you when deciding whom to vote for:
